---
title: '`ave()` for the average {dplyr} user'
description: |
  tidyverse 🤝 base R
categories:
  - dplyr
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    include-after-body: "highlighting.html"
    toc: true
    self_contained: false
    css: "../../styles.css"
editor_options: 
  chunk_output_type: console
preview: preview.png
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  R.options = list(width = 80)
)
```

I think it's safe to say that the average `{dplyr}` user does not know the `ave()` function. For that audience, this is a short appreciation post on `ave()`, a case of tidyverse _and_ base R.

## `ave()`

`ave()` is a split-apply-function in base R (specifically, `{stats}`). It's a pretty short function - maybe you can make out what it does from just reading the code^[And check out the elusive `split<-` function!]

```{r}
ave
```


Despite its (rather generic and uninformative) name, I like to think of `ave()` as actually belonging to the `*apply()` family of functions, having particularly close ties to `tapply()`.

A unique feature of `ave()` is the invariant that it **returns a vector of the same length as the input**. And if you use an aggregating function like `sum()` or `mean()`, it simply repeats those values over the observations on the basis of their grouping.

For example, whereas `tapply()` can be used to summarize the average `mpg` by `cyl`:

```{r}
tapply(mtcars$mpg, mtcars$cyl, FUN = mean)
```

The same syntax with `ave()` will repeat those values over each element of the input vector:

```{r}
ave(mtcars$mpg, mtcars$cyl, FUN = mean)
```

You can also get to this output from `tapply()` with an extra step of vectorized indexing:

```{r}
tapply(mtcars$mpg, mtcars$cyl, FUN = mean)[as.character(mtcars$cyl)]
```

## The problem

Nothing sparks more joy than when a base R function helps you write more "tidy" code. I've talked about this in length before with `outer()` in a [prior blog post on `dplyr::slice()`](https://yjunechoe.github.io/posts/2023-06-11-row-relational-operations/), and here I want to show a cool `ave()` + `dplyr::mutate()` combo.

This example is adapted from a reprex by [Cédric Scherer](https://cedricscherer.netlify.app/)^[Who I can only assume was needing this for a fancy data viz thing 😆] on the DSLC (previously R4DS) slack.

Given an input of multiple discrete columns and the frequencies of these values:

```{r}
input <- data.frame(
  a = c("A", "A", "A", "B"), 
  b = c("X", "Y", "Y", "Z"), 
  c = c("M", "N", "O", "O"), 
  freq = c(5, 12, 3, 7)
)
input
```

The task is to add new columns named `freq_*` that show the total frequency of the values in each column:

```{r}
output <- data.frame(
  a = c("A", "A", "A", "B"), 
  freq_a = c(20, 20, 20, 7),
  b = c("X", "Y", "Y", "Z"),
  freq_b = c(5, 15, 15, 7), 
  c = c("M", "N", "O", "O"), 
  freq_c = c(5, 12, 10, 10), 
  freq = c(5, 12, 3, 7)
)
output
```

So for example, in column `a` the value `"A"` is associated with values `5`, `12`, and `3` in the `freq` column, so a new `freq_a` column should be created to track their total frequencies `5 + 12 + 3` and associate that value (`20`) for all occurrences of `"A"` in the `a` column.

## Some `{tidyverse}` solutions

The gut feeling is that this seems to lack a straightforwardly "tidy" solution. I mean, the input isn't even **tidy**^[I mean that in the technical sense here. In this problem, the unit of observation is the "cells" of the input columns (the values "A", "B", "X", "Y", etc.).] in the first place!

So maybe we'd be better off starting with a pivoted tidy data for constructing a tidy solution:

```{r, message = FALSE}
library(tidyverse)
input %>% 
  pivot_longer(-freq)
```

But recall that the desired output is of a wide form like the input, so it looks like our tidy solution will require some indirection, involving something like:

```{r, eval = FALSE}
input %>% 
  pivot_longer(-freq) %>% 
  ... %>% 
  pivot_wider(...)
```

Or maybe you'd rather tackle this with some `left_join()`s, like:

```{r, eval = FALSE}
input %>% 
  left_join(summarize(input, freq_a = sum(freq), .by = a)) %>% 
  ...
```

I'll note that there's actually also an idiomatic `{dplyr}`-solution to this using the lesser-known function `add_count()`, but you can't avoid the repetitiveness problem because it doesn't vectorize on the first argument:

```{r}
input %>% 
  add_count(a, wt = freq, name = "freq_a") %>% 
  add_count(b, wt = freq, name = "freq_b") %>% 
  add_count(c, wt = freq, name = "freq_c")
```

You could try to scale this `add_count()` solution with `reduce()` (see my previous blog post on [collapsing repetitive piping](https://yjunechoe.github.io/posts/2020-12-13-collapse-repetitive-piping-with-reduce/)), but now we're straying very far from the "tidy" territory:

```{r}
input %>% 
  purrr::reduce(
    c("a", "b", "c"),
    ~ .x %>% 
      add_count(.data[[.y]], wt = freq, name = paste0("freq_", .y)),
    .init = .
  )
```

IMO this problem is actually a really good thinking exercise for the "average {dplyr} user", so I encourage you to take a stab at this yourself before proceeding if you've read this far!

## An `ave()` + `{dplyr}` solution

The crucial piece of the puzzle here is to think a little outside the box, beyond "data(frame) wrangling".

It helps to simplify the problem once we think about the problem in terms of "(column) vector wrangling" first, and that's where `ave()` comes in!

I'll start with the cake first - this is the one-liner `ave()` solution I advocated for:

```{r}
input %>% 
  mutate(across(a:c, ~ ave(freq, .x, FUN = sum), .names = "freq_{.col}"))
```

Taking column `freq_a` as an example, the `ave()` part of the solution essential creates this vector of summed-up `freq` values by the categories of `a`:

```{r}
ave(input$freq, input$a, FUN = sum)
```

From there, `across()` handles the iteration over columns and, as an added bonus, the naming of the new columns in convenient `{glue}` syntax (`"freq_{.col}"`).

It's the perfect mashup of base R + tidyverse. Base R takes care of the problem at the vector level with a split-apply-combine that's concisely expressed with `ave()`, and tidyverse scales that solution up to the dataframe level with `mutate()` and `across()`.

tidyverse 🤝 base R

## sessionInfo()

```{r}
sessionInfo()
```

