---
title: "dplyr::across(), inside out"
description: |
  Optimizing multi-column operations with type-stable functions from {collapse}
categories:
  - dplyr
  - collapse
  - data wrangling
  - tutorial
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    include-after-body: "highlighting.html"
    toc: true
    self_contained: false
    css: "../../styles.css"
editor_options: 
  chunk_output_type: console
preview: preview.png
---

```{r setup, include=FALSE}
ragg_png = function(..., res = 150) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  R.options = list(width = 80)
)
```

```{r}
library(dplyr, warn.conflicts = FALSE)
library(palmerpenguins)
```

## Big picture

- `across()` is a function that returns a data frame in the "`mutate()`" environment
- `{collapse}` has fast functions with `<data.frame>` methods which are "type-stable" (dataframe in, dataframe out)

When?

- When there exists a matrix(-like) method for the function.

- When you want to minimize the number of times the function is called ().

## `<data.frame>`s inside verbs with mutate-semantics

```{r}
toy_df <- tibble(
  id = 1:3, val = letters[1:3]
)
toy_df
```

```{r}
toy_df %>% 
  mutate( tibble(x = 1, y = 3) )
```

```{r}
toy_df %>% 
  mutate( tibble(x = 1:3, y = 3:1) )
```

## The secret life of `across()`

```{r}
penguins_length_measures <- penguins %>% 
  select( ends_with("mm") )
```

```{r}
penguins_length_measures
```

```{r}
penguins %>% 
  transmute( across(.cols = ends_with("mm")) )
  # mutate( across(where(is.numeric)), .keep = "none" )
```

```{r}
penguins %>% 
  transmute( penguins_length_measures )
```


## Case #1 - scale

```{r}
penguins %>% 
  transmute( across(ends_with("mm"), scale) )
```

```{r}
penguins %>% 
  transmute( across(ends_with("mm"), ~ scale(.x)[,1]) )
```

BUT

```{r}
penguins %>% 
  transmute(
    across(ends_with("mm")) %>% 
      scale() %>% 
      as.data.frame()
  )
```

## Case #2 - "cell"-operations

```{r}
penguins %>% 
  transmute( across(ends_with("mm"), ~ .x * 10) )
```

```{r}
library(stringr)
penguins %>% 
  transmute( across(ends_with("mm"), ~ .x * 10, .names = "{str_replace(.col, '_mm', '_cm')}") )
```

```{r}
penguins %>% 
  transmute(
    across(ends_with("mm")) %>% 
      as.matrix() %>% 
      `*`(10) %>% 
      as.data.frame() %>% 
      rename_with(~ str_replace(.x, '_mm', '_cm'))
  )
```

```{r}
library(purrr)
penguins %>% 
  transmute(
    across(ends_with("mm")) %>% 
      purrr::modify(~ .x * 10) %>% 
      rename_with(~ str_replace(.x, '_mm', '_cm'))
  )
```

```{r}
library(purrr)
penguins %>% 
  transmute(
    across(ends_with("mm"), .names = "{str_replace(.col, '_mm', '_cm')}") %>% 
      purrr::modify(~ .x * 10)
  )
```

## Case #... - rowsums

```{r}
df <- data.frame(
    Type1_A = 1:10,
    Type1_B = 11:20,
    Type2_A = 1:10,
    Type2_B = 11:20
)
df %>% 
  mutate(
    A = rowSums( across(ends_with("_A")) ),
    B = rowSums( across(ends_with("_B")) )
  )
```


## Case #3 - Vectorized matrix functions (relationships between columns)

```{r}
library(tidyr)
billboard %>% 
  filter( if_any(starts_with("wk"), ~ .x == 1) ) %>% 
  select(artist, track)
```

```{r}
billboard %>% 
  filter( do.call(pmin, c(across(starts_with("wk")), na.rm = TRUE)) == 1 ) %>% 
  select(artist, track)
```

```{r}
billboard %>% 
  filter( do.call(pmin, c(across(starts_with("wk")), na.rm = TRUE)) == 1 ) %>% 
  transmute(
    artist, track,
    top_weeks = across(starts_with("wk")) %>% 
      as.matrix() %>% 
      `==`(1) %>% 
      rowSums(na.rm = TRUE)
  )
```


## Powerup with collapse

## Collapse for #1

```{r}
library(collapse)
```

```{r}
penguins %>% 
  transmute(
    fscale( across(ends_with("mm")) )
  )
```

## Collapse for #2

```{r}
penguins %>% 
  transmute(
    across(ends_with("mm"), .names = "{str_replace(.col, '_mm', '_cm')}") #??
  )
```


## Collapse for #3

```{r}
billboard %>% 
  filter( browser() ) %>% 
  select(artist, track)
```


## Collapse complex

```{r}
penguins %>% 
  group_by(species) %>% 
  transmute( fscale(across(ends_with("mm"))) ) %>% 
  ungroup()
```

```{r}
penguins %>% 
  transmute( fscale(across(ends_with("mm")), species) ) # `species` can be sym!
```

```{r}
fscale(
  x = penguins %>% select(ends_with("mm")),
  g = penguins$species
)
```

```{r}
penguins %>% 
  transmute(
    # grouped by species
    fscale(across(ends_with("mm"), .names = "{str_remove(.col, '_mm$')}"), species),
    # grouped by island
    body_mass_g = fscale(body_mass_g, island),
    # other ungrouped operations
    species = toupper(species),
    island = tolower(island)
  )
```

### Whew

```{r}
library(dplyr)
library(collapse)
mtcars
```

