---
title: "Saving a line of piping"
description: |
  Some notes on lesser known functions/functionalities that combine common chain of {dplyr} verbs.
categories:
  - data wrangling
  - dplyr
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
preview: preview.png
date: 08-07-2020
output:
  distill::distill_article:
    toc: true
    self_contained: false
---

```{r opts}
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  R.options = list(width = 80)
)
```

```{r, message = FALSE, echo = FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
options(dplyr.summarise.inform = FALSE)
```

Using a cleaned up version of `penguins` data from `{palmerpenguins}`:

```{r}
data("penguins", package = "palmerpenguins")

penguins <- na.omit(penguins)
```

```{r, layout="l-body-outset", echo = FALSE}
reactable::reactable(penguins, defaultPageSize = 5,
                     wrap = FALSE, bordered = TRUE, sortable = FALSE)
```

## 1. `rename()` inside `select()`

You can rename a column inside `select()` by assigning a new name on the left hand side:

```{r}
##### Long Form #####
# penguins %>% 
#   select(species, island) %>% 
#   rename(penguin_species = species)

penguins %>% 
  select(penguin_species = species,
         island)
```

This also works with `{tidyselect}` helpers like `starts_with()`, `ends_with()`, `contains()`, and `matches()`:

```{r}
##### Long Form #####
# penguins %>% 
#   select(species, island) %>% 
#   rename(penguin_species = species,
#          weight = body_weight_g)

penguins %>% 
  select(penguin_species = species,
         island,
         weight = contains("mass"))
```


## 2. `rename()` inside `count()`

You can rename the new column of counts (`n` by default) using the `name` argument:

```{r}
##### Long Form #####
# penguins %>% 
#   count(species) %>% 
#   rename(total = n)

penguins %>% 
  count(species, name = "total")
```

You can also rename the column(s) that are selected for counting in the same way as shown in the `select()` examples above:

```{r}
##### Long Form #####
# penguins %>% 
#   count(species) %>% 
#   rename(total = n,
#          penguin_species = species)

penguins %>% 
  count(penguin_species = species, name = "total")
```

Note that the new name passed into the `name` argument must be quoted, but the new name for selected column needs not to be unquoted:

```{r}
identical(
  # Method 1: new column name UNQUOTED
  penguins %>% 
    count(penguin_species = species, name = "total"),
  # Method 2: new column name QUOTED
  penguins %>% 
    count("penguin_species" = species, name = "total") 
)
```

I prefer to unquote the new column names to keep it consistent with the recommended style for `rename()`

This feature of `select()` may seem weird and hackish (and I guess it sort of is in this demonstration) but it's explicitly documented [here](https://tidyselect.r-lib.org/articles/syntax.html#renaming-variables-1) if you want to read more on it.

## 3. `mutate()` inside `count()`

You can also create a new column to count by inside `count()`. This works very similarly to the above, but I think it's worth its own mention.

It's pretty simple - you just do what you'd do for `mutate()` inside `count()`:

```{r}
##### Long Form #####
# penguins %>% 
#   mutate(long_beak = bill_length_mm > 50) %>% 
#   count(long_beak)

penguins %>% 
  count(long_beak = bill_length_mm > 50)
```

And of course, this also works when specifying _multiple_ variables to count by:

```{r}
##### Long Form #####
# penguins %>% 
#   mutate(long_beak = bill_length_mm > 50,
#          is_adelie = species == "Adelie") %>% 
#   count(is_adelie, long_beak)

penguins %>% 
  count(long_beak = bill_length_mm > 50,
        is_adelie = species == "Adelie")
```

## 4. `transmute()` + `select()`

`transmute()` is a function that mutates columns and returns only those columns:

```{r}
##### Long Form #####
# penguins %>% 
#   mutate(body_mass_kg = body_mass_g/1000) %>% 
#   select(body_mass_kg)

penguins %>% 
  transmute(body_mass_kg = body_mass_g/1000)
```

I've rarely used `transmute()` in the past because I thought it could only return _modified_ columns, which would be very limiting (like in the above example, what good is a single column of penguin body mass in kilograms?)

But actually you can just name the columns you want to include in `transmute()` like you would in `select()` to carry over columns that you aren't modifying. And of course, you can "rename" them as you do it^[What happens under the hood is actually copying of a sort, so this is probably not the best approach if you care about efficiency. As a case in point, you can't use `{tidyselect}` helpers in `transmute` because you're creating a new dataframe]:

```{r}
##### Long Form #####
# penguins %>% 
#   mutate(body_mass_kg = body_mass_g/1000) %>% 
#   select(species, island, body_mass_kg) %>% 
#   rename(penguin_species = species)

penguins %>% 
  transmute(penguin_species = species,
            island,
            body_mass_kg = body_mass_g/1000)
```


## 5. `ungroup()` inside `summarize()`

I always found using `ungroup()` after `summarize()` to be _extremely_ ugly, but I found myself using it a lot to remove left-over groupings after a `summarize` call:

```{r}
penguins %>% 
  group_by(island, species) %>% 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) %>% 
  ungroup()
```

... because `summarize()` only drops the last grouping variable by defaut, meaning that the output is still grouped by the `island` variable if `ungroup()` isn't called:

```{r}
# Without ungroup()
penguins %>% 
  group_by(island, species) %>% 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) %>% 
  group_vars()

# With ungroup()
penguins %>% 
  group_by(island, species) %>% 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE)) %>% 
  ungroup() %>% 
  group_vars()
```

Since `{dplyr}` 1.0.0, you can simply set the `.groups` argument inside `summarize()` to `'drop'` to achieve the same:

```{r}
penguins %>% 
  group_by(island, species) %>% 
  summarize(mean_mass = mean(body_mass_g, na.rm = TRUE), .groups = 'drop')
```

But `ungroup()` still remains relevant as you can now selectively remove grouping variables in {dplyr} 1.0.0.


## 6. `arrange()` + other features inside `slice()`

In past versions of `{dplyr}`, if you wanted to grab the top n rows sorted by a column, you'd use `top_n()`, which provides a simpler way of doing `slice()` + `arrange()`:

```{r}
##### Long Form #####
# penguins %>% 
#   arrange(desc(body_mass_g)) %>% 
#   slice(1:5)

penguins %>% 
  top_n(5, wt = body_mass_g)
```

But the recent `{dplyr}` 1.0.0 augmented `slice()` with variants like `slice_min()` and `slice_max()` that now supresede `top_n()`:

```{r}
##### Pre-1.0.0 #####
# penguins %>% 
#   top_n(5, wt = body_mass_g)

penguins %>% 
  slice_max(order_by = body_mass_g, n = 5)
```

Note that the order of arguments is different for `slice_min/max()` - the first argument after piping is where you specify the variable for ordering rather than the number of rows to return, like in `top_n()`.

This is because `slice_min/max()` gives you an option to either specify a certain number of rows `n` or a proportion of rows `prop`:

```{r}
penguins %>% 
  slice_max(body_mass_g, prop = .01)
```

And actually, the most significant change with the new `slice_*()` functions is from adding appropriate behavior for **grouped dataframes**.

So for example, this example below returns the top 5% of penguins by weight for each species:

```{r}
penguins %>% 
  group_by(species) %>% 
  slice_max(body_mass_g, prop = .05)
```

But note that `slice_*()` functions do not modify groups in the result if the input is a grouped dataframe, so you need to explicitly add a call to `ungroup()` if you want to drop groups after slicing.

## 7. count and sum by group with `add_count()`

Saving my favorite lesser-known `{dplyr}` function for last!

`add_count()` adds a column with the counts of each group (or combination of groups):

```{r}
##### Long Form #####
# penguins %>% 
#   group_by(species) %>% 
#   mutate(count_by_species = n()) %>% 
#   ungroup()

penguins %>% 
  add_count(species, name = "count_by_species") %>% 
  # cutting down some columns to show the new column
  select(-contains("mm"))
```

You can use the `wt` to effectively get sums by group (perhaps hackish but very very useful):

```{r}
##### Long Form #####
# penguins %>% 
#   group_by(species) %>% 
#   mutate(total_weight_by_species = sum(body_mass_g)) %>% 
#   ungroup()
  

penguins %>% 
  add_count(species, wt = body_mass_g, name = "total_weight_by_species") %>% 
    # cutting down some columns to show the new column
  select(-contains("mm"))
```

Also check out its more primitive version `add_tally()`.

By default, `add_tally()` adds a count of rows, which you can already do with `mutate(n = n())`, but it shines when you make use of its `wt` argument:

```{r}
penguins %>% 
  add_count(species, wt = body_mass_g, name = "total_weight_by_species") %>% 
  add_tally(wt = body_mass_g, name = "total_weight_of_all_species") %>% 
  select(1:2, last_col(0):last_col(1))
```