---
title: Fonts for graphs
description: |
  A small collection of my favorite fonts for data visualization
categories:
  - data visualization
  - typography
base_url: https://yjunechoe.github.io
preview: preview.png
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: 09-06-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r opts}
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  R.options = list(width = 80)
)
```

&nbsp;

For the last few weeks I've been reading about and experimenting with fonts for data visualization in my spare time.^[I reference a lot this great collection of fonts used in profesional visualization [here](https://fontsinuse.com/).] Out of that, I have found a couple fonts that I really like and wanted to do a small showcase of them here.

These fonts are all **free and available for download** at [Google Fonts](https://fonts.google.com/).^[For how to import local fonts into R to use for plotting, check out {extrafont} and/or {showtext}.] Note that not only are they all large font families that come with many different styles, you can also adjust various theme settings like `lineheight` in {ggplot2}, so what I'm showing here isn't the full extent of what you can make with these fonts.

&nbsp;


```{r code, results = 'asis', echo = FALSE, message = FALSE}
library(ggplot2)
library(extrafont)
knitr::opts_chunk$set(fig.width = 7, dpi = 600)

theme_set(theme_classic(base_size = 14))

favorites <- c(
  "Arial" = "It's the default. It's dull. It's just here for comparison.",
  "Montserrat" = "Simple design that can handle long lines of text. I like it for minimal plots.",
  "Roboto Mono" = "Monospaced member of the Roboto family. Very easy to read.",
  "Futura Bk BT" = "A slender and bold member of the Futura family. Looks nice even in larger sizes.",
  "Barlow" = "Also a slender font like Futura, but this has nicer 'j's",
  "Adelle" = "A serif font that doesn't go overboard. I use it a lot for short paragraphs.",
  "Merriweather" = "Similar to Adelle, but has a bit more pronounced hooks"
)


for (font in names(favorites)) {
  cat("\n\n## ", font, "\n\n")
  cat("", favorites[font], "\n\n")
  plot <- qplot(data = mtcars, mpg, disp, color = factor(cyl)) +
    annotate("text", 28, 400, label = paste(letters, collapse = ''), family = font) +
    geom_curve(aes(x = 28, y = 380, xend = 22, yend = 260),
               color = 'black', curvature = -.3, arrow = arrow(), show.legend = FALSE) +
    labs(title = "This is an interesting plot title",
         subtitle = "Here's the subtitle 1234567890",
         caption = "This is the plot caption") +
    theme(text = element_text(family = font),
          plot.title.position = 'plot')
  print(plot)
}
```



&nbsp;

# **Misc.**

Why spend 3 minutes copy-pasting code when you can spend an hour automatizing it?

This was my first time using dynamic Rmarkdown reporting. The plots above and the text descriptions that went with them were generated in a `for` loop, which I learned about [here](https://bookdown.org/yihui/rmarkdown-cookbook/reuse-chunks.html). 

Here is the single chunk of code that made this possible:

```{r, eval = FALSE}
<<code>>
```