---
title: "Vector construction with matrices"
description: |
  Kidnapping multidimensional space
categories:
  - data visualization
  - ggplot2
  - tutorial
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    self_contained: false
    css: "../../styles.css"
editor_options: 
  chunk_output_type: console
draft: true
---

```{r setup, include=FALSE}
library(knitr)
ragg_png = function(..., res = 150) {
  ragg::agg_png(..., res = res, units = "in")
}
opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  R.options = list(width = 80)
)
```

## Introduction

Generate number sequences

Sequentially:

```{r}
rep(1:3, 3)
```

Algebraically:

```{r}
0:8 %% 3 + 1
```

Geometrically:

```{r}
as.integer( matrix(rep(1:3, 3), ncol = 3) )
```

When sequences are understood as the "interweaving of multiple vectors":

```{r}
c(1, 3, 7, 2, 4, 8, 3, 6, 9)
```

Which contains members from three vectors:

```{r}
c(1, 2, 3)
c(3, 4, 5)
c(7, 8, 9)
```

This looks oddly like a matrix:

```{r}
matrix(1:9, ncol = 3)
```

The primary axis of matrices are columns (much like dataframes), so when we "collapse" a matrix into a vector, you "collect" the cell values starting from the top-left cell and zig zag top-to-bottom, left-to-right until you hit the bottom-right cell. 

```{r}
as.integer( matrix(1:9, ncol = 3) )
```

Exploiting this fact, it'd be nice if we could generate our number sequence via the use of a matrix:

```{r}
matrix(c(1, 3, 7, 2, 4, 8, 3, 6, 9), ncol = 3)
```

You'd be surprised to know that this is just our boring matrix that has been rotated:

```{r}
t( matrix(1:9, ncol = 3) )
```

We get our complex sequence if we collapse this rotated matrix:

```{r}
as.integer( t( matrix(1:9, ncol = 3) ) )
```

What are some other fun things we can do with matrices?

```{r}
library(dplyr)

mtcars %>% 
  mutate(across(vs:carb, identity, .names = "{.col}_copy"))
```


