---
title: Shiny tips - the first set
description: |
  %||%, imap() + {shinybusy}, and user inputs in modalDialog()
categories:
  - shiny
base_url: https://yjunechoe.github.io
preview: preview.png
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: 07-20-2020
output:
  distill::distill_article:
    self_contained: false
---

```{r opts}
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  R.options = list(width = 80)
)
```

When I was an RA in the [LEARN lab](https://learnlab.northwestern.edu/) - a child language development lab at Northwestern - I worked on a shiny app that automates snowball search for meta-analysis research ([relevant research poster](../../../static/poster/Snowballer_June_Choe.pdf)). Long story short, I worked on it for a couple months, got it working, then stopped working on it for another couple months, and had the chance to revisit it just recently.

When I picked the project back up, I realized that my old code was poorly commented, somewhat inefficient, and even hackish at times. So I decided to re-write it from scratch. In this second time around, I learned a lot of useful functions/tricks that really helped streamline my code and I thought I'd document my three favorite ones here for future reference.

### 1. `%||%` from `{rlang}`

Basically, `%||%` is an infix operator that returns the left-hand side when it is not `Null` and the right-hand side when it is `Null`. It's from the `{rlang}` package but you can also define the function yourself:

```{r}
"%||%" <- function(lhs, rhs) {
  if (is.null(lhs)) rhs else lhs
}

a <- 10
b <- NULL

a %||% b
b %||% a

# note how the output is different when b is no longer null
b <- 11
b %||% a
```

I found this operator to be extremely useful when displaying empty tables as placeholders when using `DT::datatable()`. It allows me to communicate to the user where a table is expected to appear rather than just not showing anything at all when no data is loaded (which is what happens by default).

For example, if you want to show an empty column with an empty row when the data (here, the reactive variable `mydf`) is null, you might do the following:

```{r, eval = FALSE}
mydf_display <- renderDataTable({
  datatable(mydf() %||% tibble(` ` = NA))
})
```

Another use-case for `%||%` is when I'm trying a sequence of function calls until one one of them succeeds and returns a non-null value. For example, say I want to scrape some information online and I have API wrappers for different websites that potentially have that information. I can chain them together using `%||%` like so:

```{r, eval = FALSE}
myinfo <- 
  scrape_website1() %||%
  scrape_website2() %||%
  scrape_website3()
```

This is much neater than nested if...else statements!


### 2. `purrr::imap()` and `{shinybusy}`

Using my shiny app involves a lot of waiting (querying online databases), so I looked into ways to show a progress bar similar to the family of `*Modal()` functions from `{shiny}`. The extension package `{shinybusy}` ([project site](https://dreamrs.github.io/shinybusy/index.html)) offers a very satisfying solution to this problem.

Basically, you initialize a progress bar with `show_modal_progress_*()` and increment its value inside whatever operation you're doing. Here's a pseudo code demonstrating how it works:

```{r, eval = FALSE}
initialize a progress bar

create a new_list of same size to store output

for index in seq_along(list):
  new_list[index] <- calculations(list[index])
  increment progress bar by index
  
remove progress bar

return new_list
```

But in my case, my "do stuff" part didn't involve a big wall of code because I packed it into a single function in a separate file that I source at the beginning. This, coupled with my general aversion to for-loops, drove me to `imap()` and its variants from `{purrr}`. `imap()` is like `map()` except it also keeps track of the index of the element that you're operating on (to put it another way, it's like `map2()` where `.y` is the index).

Now, you don't need an explicit for-loop to increment and the above code can be reduced to this:

```{r, eval = FALSE}
initialize a progress bar

new_list <- imap(list,
                 ~{
                   calculations(.x)
                   increment progress bar by .y
                 })
  
remove progress bar

return new_list
```

In my opinion, this is much cleaner! For a more concrete example, here's a template using actual code:

```{r, eval = FALSE}
my_data <- eventReactive(input$my_button, {
  
  # initialize a progress bar
  show_modal_progress_line()
  
  # do operation on elements of vector
  result <- imap(my_reactive_var(),
                 ~{
                   update_modal_progress(value = .y / length(my_reactive_var()))
                   my_operation_on_element(.x)
                 })
  
  # remove progress bar
  remove_modal_progress()
  
  # return output
  return(result)
  
})
```



### 3. User inputs inside `modalDialog()`

In `{shiny}`, you can show the user a pop-up message box by first laying out the content of the message in `modalDialog()` and then rendering it with `showModal()`. In the first version of my app, I used this to show simple messages like warnings, but did you know that you can include any `*Input` widgets too?

For example, this code renders a pop-up box for a file upload in response to a button click:

```{r, eval = FALSE}
observeEvent(input$MyButton, {
  showModal(modalDialogue(
    title = "Upload File Here",
    fileInput(inputID = "UploadedFile", label = "Upload")
  ))
})
```

And you can access whatever is uploaded using `input$UploadedFile` like you would if the file upload widget was in the `ui` side of the app!

This took me a bit to get used to because you are defining the modal in the `server` side where the content of the modal looks like the `ui` side but can be accessed back at the `server` side. But this was life-changing and it opened up a lot of potential for my GUI to be less cluttered. Using this neat trick, I was able to move a large feature into a modal that would only be available upon a click of a button (it was a feature designed for a rare case scenario so I thought I'd save the user from having to see the entire interface for that if they don't ask for it).


### Ending note

The more I learn and use shiny, the less I feel like I know. I'm actually enjoying this stage of my progress because every new thing just absolutely wows me (and I hope to continue sharing what I learn - hence this being the "first set"). And very much looking forward to [Hadley Wickham's new book on shiny](https://mastering-shiny.org/)!






