---
title: 'args(args(args)(args))'
description: |
  The unexpected sequal to "R is a language optimized for meme-ing"
categories:
  - args
  - metaprogramming
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    include-after-body: "highlighting.html"
    toc: true
    self_contained: false
    css: "../../styles.css"
editor_options: 
  chunk_output_type: console
preview: preview.png
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  R.options = list(width = 80)
)
```

The kind of blog posts that I have the most fun writing are those where I hyperfocus on a single function, like [dplyr::slice()](https://yjunechoe.github.io/posts/2023-06-11-row-relational-operations/), [purrr::reduce()](https://yjunechoe.github.io/posts/2020-12-13-collapse-repetitive-piping-with-reduce/), and [ggplot2::stat_summary()](https://yjunechoe.github.io/posts/2020-09-26-demystifying-stat-layers-ggplot2/). In writing blog posts of this kind, I naturally come across a point where I need to introduce the argument(s) that the function takes. I usually talk about them one at a time as needed, but I *could* start by front-loading that important piece of information first.

In fact, there's a function in R that lets me do exactly that, called `args()`.

## `args()`

`args()` is, in theory, a very neat function. According to `?args`:

> Displays the argument names and corresponding default values of a (non-primitive or primitive) function.

So, for example, I know that `sum()` takes the arguments `...` and `na.rm` (with the `na.rm = FALSE` default). The role of `args()` is to display exactly that piece of information using R code. This blog runs on rmarkdown, so surely I can use `args()` as a convenient and fancy way of showing information about a function's arguments to my readers.

In this blog post, I want to talk about `args()`. So let's start by looking at the argument that `args()` takes.

Of course, I could just print `args` in the console:

```{r}
args
```

But wouldn't it be fun if I used `args()` itself to get this information?

## `args(args)`

```{r}
args(args)
```

Okay, so I get the `function (name)` piece, which is the information I wanted to show. We can see that `args()` takes one argument, called `name`, with no defaults.

But wait - what's that `NULL` doing there in the second line?

Hmm, I wonder if they forgot to `invisible()`-y return the `NULL`. `args()` is a function for *displaying* a function's arguments after all, so maybe the arguments are printed to the console as a side-effect and the actual output of `args()` is `NULL`.

If that is true, we should be able to suppress the printing of `NULL` with `invisible()`:

```{r}
invisible(args(args))
```

Uh oh, now *everything* is invisible.

Alright, enough games! What exactly are you, output of `args()`?!

```{r}
typeof(args(args))
```

What? 

## `args(args)(args)`

Turns out that `args(args)` is actually returning a whole function that's a *copy* of `args()`, except with its body replaced with NULL.

So `args(args)` is itself a function that takes an argument called `name` and then returns `NULL`. Let's assign it to a variable and call it like a function:

```{r}
abomination <- args(args)
```

```{r}
abomination(123)
abomination(mtcars)
abomination(stop())
```

The body is *just* `NULL`, so the function doesn't care what it receives^[You can even see lazy evaluation in action when it receives `stop()` without erroring.] - it just returns `NULL`.

In fact, we could even pass it... `args`:

```{r}
args(args)(args)
```

## `args(args(args)(args))`

But wait, that's not all! `args()` doesn't *just* accept a function as its argument. From the documentation:

> **Value**
>
> NULL in case of a non-function.

So yeah - if `args()` receives a non-function, it just returns `NULL`:

```{r}
args(123)
args(mtcars)
```

This applies to *any* non-function, including... `NULL`:

```{r}
args(NULL)
```

And recall that:

```{r}
is.null( args(args)(args) )
```

Therefore, this is a valid expression in base R:

```{r}
args(args(args)(args))
```

## ad infinitum

For our cursed usecase of using `args(f)` to return a copy of `f` with it's body replaced with `NULL` only to then immediately call `args(f)(f)` to return `NULL`, it really doesn't matter what the identity of `f` is as long as it's a function.

That function can even be ... `args(args)`!

So let's take our `args(args(args)(args))`:

```{r}
args( args( args )( args ))
```

And swap every `args()` with `args(args)`:

```{r}
args(args)( args(args)( args(args) )( args(args) ))
```

Or better yet, swap every `args()` with `args(args(args))`:

```{r}
args(args(args))( args(args(args))( args(args(args)) )( args(args(args)) ))
```

The above unhinged examples are a product of two patterns:

1) The fact that you always get `function (name) NULL` from wrapping `args()`s over `args`:

    ```{r}
    list(
      args(          args),
      args(     args(args)),
      args(args(args(args)))
    )
    ```

2) The fact that you can get this whole thing to return `NULL` by having `function (name) NULL` call the function object `args`. You can do this anywhere in the stack and the `NULL` will simply propagate:

    ```{r}
    list(
      args(args(args(args))) (args)   ,
      args(args(args(args))  (args) ) ,
      args(args(args(args)   (args) ))
    )
    ```

We could keep going but it's tiring to type out and read all these nested `args()`... but did you know that there's this thing called the pipe `%>%` that's the solution to all code readability issues?

## Had enough `args()` yet?

Let's make an `args()` factory `ARGS()` ...

```{r}
library(magrittr)
ARGS <- function(n) {
  Reduce(
    f = \(x,y) bquote(.(x) %>% args()),
    x = seq_len(n),
    init = quote(args)
  )
}
```

... to produce a sequence of `args()` ...

```{r}
ARGS(10)
eval(ARGS(10))
```

... and tidy it up!

```{r}
ARGS(10) %>% 
  deparse1() %>% 
  styler::style_text()
```

Wanna see even more unhinged?

Let's try to produce a "matrix" of `args()`. You get a choice of `i` "rows" of piped lines, and `j` "columns" of `args()`-around-`args` each time - all to produce a `NULL`.

Ready?

```{r}
ARGS2 <- function(i, j) {
  Reduce(
    f = \(x,y) bquote(.(x) %>% (.(y))),
    x = rep(list(Reduce(\(x,y) call("args", x), seq_len(j), quote(args))), i)
  )
}
```

```{r}
ARGS2(5, 1) %>% 
  deparse1() %>%
  styler::style_text()
```

```{r}
ARGS2(5, 3) %>% 
  deparse1() %>%
  styler::style_text()
```

```{r}
ARGS2(10, 5) %>% 
  deparse1() %>%
  styler::style_text()
```

```{r}
list(
  eval(ARGS2(5, 1)),
  eval(ARGS2(5, 3)),
  eval(ARGS2(10, 5))
)
```

Yay!

## TL;DR: `str()`

If you want a version of `args()` that does what it's supposed to, use `str()` instead:^[Though you have to remove the `"srcref"` attribute if the function has one. But also don't actually do this!]

```{r}
str(args)
str(sum)
```

`args()` is hereafter banned from my blog.

## Coda (serious): redesigning `args()`

The context for my absurd rant above is that [I was just complaining](https://fosstodon.org/@yjunechoe/112039945400602627) about how I think `args()` is a rather poorly designed function.

Let's try to redesign args. I'll do three takes:

### Take 1) Display is the side-effect; output is trivial

If the whole point of `args()` is to **display** a function's arguments for inspection in interactive usage, then that can simply be done as a side-effect.

As I said above, `str()` surprisingly has this more sensible behavior out of the box. So let's write our first redesign of `args()` which just calls `str()`:

```{r}
args1 <- function(name) {
  str(name)
}
args1(sum)
```

In `args1()`/`str()`, information about the function arguments are sent to the console.^[Technically, the `"output"` stream.] We know this because we can't suppress this with `invisible` but we *can* grab this via `capture.output`:

```{r}
invisible( args1(sum) )
capture.output( args1(sum) )
```

For functions whose purpose is to signal information to the console (and whose usage is limited to interactive contexts), we don't particularly care about the output. In fact, because the focus isn't on the output, the return value should be as *trivial* as possible.

A [recommended option](https://design.tidyverse.org/out-invisible.html) is to just invisibly return `NULL`. This is now how `args1()` does it (via `str()`).^[For the longest time, I thought `args()` was doing this from how its output looked.]:

```{r}
print( args1(sum) )
is.null( args1(sum) )
```

Alternatively, the function could just invisibly return what it receives,^[Essentially acting like `identity()`.] which is another common pattern for cases like this. Again, we return invisibly to avoid distracting from the fact that the point of the function is to *display* as the side-effect.

```{r}
args2 <- function(name) {
  str(sum)
  invisible(name)
}
```

```{r}
args2(rnorm)
```

```{r}
args2(rnorm)(5)
```


### Take 1) Display is the side-effect; output is meaningful

One thing I neglected to mention in this blog post is that there are other ways to extract a function's arguments. One of them is `formals()`:^[But note that it has a [special behavior](https://adv-r.hadley.nz/functions.html?q=formals#primitive-functions) of returning `NULL` for primitive functions (written in C) that clearly have user-facing arguments on the R side. See also `formalArgs()`, for a shortcut to `names(formals())`]

```{r}
formals(args)
formals(rnorm)
```

`formals()` returns the information about a function's arguments in a list which is pretty boring, but it's an object we can manipulate (unlike the return value of `str()`). So there's some pros and cons.

Actually, we could just combine both `formals()` and `str()`:

```{r}
args3 <- function(name) {
  str(name)
  invisible(formals(name))
}
```


```{r}
arguments <- args3(rnorm)
arguments
arguments$mean
```

You get the nice display as a side-effect (via `str()`) and then an informative output (via `formals()`). You could even turn this into a class with a print method, which is definitely the better way to go about this, but I'm running out of steam here and I don't like OOP, so I won't touch that here.

### Take 3) Just remove the `NULL`

This last redesign is the simplest of the three, and narrowly deals with the problem of that pesky `NULL` shown alongside the function arguments:

```{r}
args(sum)
```

Fine, I'll give them that `args()` must, for ~~compatibility with S~~ whatever reason, return a whole new function object, which in turn requires a function body. But if that function is just as a placeholder and not meant to be called, can't you just make the function body, like, empty?

```{r}
args4 <- function(name) {
  f <- args(name)
  body(f) <- quote(expr=)
  f
}
args4(sum)
args4(rnorm)
typeof( args4(rnorm) )
```

Like, come on!
