---
title: '{ggplot2} metaprogramming patterns'
description: |
  A ggplot blog post that only uses `aes()`
categories:
  - ggplot2
  - metaprogramming
base_url: https://yjunechoe.github.io
author:
  - name: June Choe
    affiliation: University of Pennsylvania Linguistics
    affiliation_url: https://live-sas-www-ling.pantheon.sas.upenn.edu/
    orcid_id: 0000-0002-0701-921X
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    include-after-body: "highlighting.html"
    toc: true
    self_contained: false
    css: "../../styles.css"
editor_options: 
  chunk_output_type: console
preview: preview.png
---

```{r setup, include=FALSE}
library(ggplot2)
knitr::opts_chunk$set(
  comment = " ",
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  R.options = list(width = 80)
)
```

## ggplot2 metaprogramming

In `{ggplot2}`, aesthetic mappings are declared using `aes()`. If you want to plot a time series of daily average temperature from a data where the `day` column is mapped to the x-axis and the `temperature` column is mapped to the y-axis, you'd write something like:

```{r, eval=FALSE}
aes(day, temperature)
```

Or more explicitly,

```{r, eval=FALSE}
aes(x = day, y = temperature)
```

When we write ggplot code, we don't really do much with the `aes()` function alone. I do this when I'm teaching ggplot too - for the sake of simplicity, I actively try not to draw attention to the fact that `aes()` is itself a function. I just tell my students that the `aes()` is "a place where you write down the aesthetic mappings", and that simply mental model can get users very far.

But this is a deceptively simple understanding of `aes()` - one that we have to unlearn when we start doing more advanced stuff, like writing *functions* that return ggplot objects.

The world of `aes()` will look simultaneously familiar, yet at times overwhelmingly foreign. This blog post will try to showcase a little bit of that in (hopefully) a gentle way.

We start with the most obvious yet most under-appreciated fact: `aes()` returns a stand-alone object.

```{r}
x <- aes(day, temperature)
x
```

## The structure of `aes()`

`aes()` returns an object of class `<uneval>`:

```{r}
class(x)
```

It means "unevaluated expression(s)". It's unevalated because it "captures" (a.k.a. ["defuses"](https://rlang.r-lib.org/reference/topic-defuse.html)) what we, as the user, provided to `aes()`.

We know `time` and `temperature` are unevaluated by `aes()` because if we _were_ to evaluate them, it would error. And of course they do - they're undefined variables!

```{r, error=TRUE}
day
temperature
```

This `<uneval>` object returned by `aes()` is actually just a list: 

```{r}
typeof(x)
```

And you can see its list-like nature when you strip away its class:

```{r}
unclass(x)
```

## sessionInfo()

```{r}
sessionInfo()
```

